---
title: "Relationships and multivariate methods"
subtitle: "Block 5 · exercises 6 – 9"
---

```{r}
#| include: false
source("../setup.R")
library(MASS)
sensory <- read.table("http://senzorika.com/sait/datasety/senzorika.txt", sep = ",")
```

## Correlation: direction and strength

```{r}
#| fig-height: 3.2
set.seed(3); x <- rnorm(40)
par(mfrow = c(1, 3))
for (k in list(c(0.9, 0.4), c(0, 1), c(-0.8, 0.55))) {
  y <- k[1] * x + rnorm(40) * k[2]
  plot(x, y, pch = 19, col = pal[3], main = paste("r =", round(cor(x, y), 2)), xlab = "", ylab = "")
}
```

::: small
Pearson – a linear relationship. **Spearman** – a monotonic relationship on ranks (suitable for sensory scales).
:::

## Correlation is not everything

```{r}
#| output-location: column
sugar <- 1:13
pleasantness <- c(5, 5, 5, 7, 9, 9, 9, 8, 9, 7, 7, 6, 5)
plot(sugar, pleasantness, type = "b", pch = 19,
     col = pal[2], lwd = 2)
cor(sugar, pleasantness, method = "spearman")
```

::: warn
The correlation is almost zero although the relationship is strong – it has an **optimum**. Always plot first.
:::

## Simple linear regression

```{r}
#| output-location: column
sweetness <- c(2, 2, 3, 4, 5, 6, 6, 6, 7, 9, 9, 9, 8)
model <- lm(sweetness ~ sugar)
plot(sugar, sweetness, pch = 19, col = pal[3])
abline(model, col = pal[1], lwd = 3)
segments(sugar, sweetness, sugar, fitted(model),
         col = pal[2])
round(coef(model), 2)
summary(model)$r.squared
```

## PCA – many attributes, two axes

```{r}
#| output-location: column
survey <- data.frame(
  sensory = c(3,7,7,5,5,6,5,6,3,3,2,5,2,7,1,2),
  price = c(5,3,4,7,7,4,7,5,5,3,6,7,4,5,6,3),
  design = c(3,2,4,1,5,4,2,4,6,5,6,7,5,5,5,7),
  brand = c(7,3,3,3,5,3,3,4,3,5,3,3,6,5,5,7))
pca <- princomp(survey, cor = TRUE)
round(pca$sdev^2, 2)            # eigenvalues
biplot(pca, col = c("grey40", pal[2]), cex = 0.8)
```

::: small
Kaiser: keep components with an eigenvalue &gt; 1. Comp.1 sets **taste** against **image** (brand, package).
:::

## Factor analysis

```{r}
fa_data <- data.frame(
  quality = c(5,3,4,7,7,4,7,5,5,3,6,7,4,5,6,3), price = c(6,7,6,5,7,6,5,6,3,1,2,5,2,3,1,2),
  aesthetics = c(3,2,4,1,5,2,2,4,6,7,6,7,5,6,5,7), brand = c(4,2,5,3,5,3,1,4,7,5,7,6,6,5,5,7),
  friend = c(7,2,5,6,2,4,1,7,3,2,6,7,6,2,4,5), family = c(6,3,4,7,1,5,4,5,4,4,5,7,2,3,5,6))
round(eigen(cor(fa_data))$values, 2)
print(factanal(fa_data, factors = 2)$loadings, cutoff = 0.4)
```

## Cluster analysis: dendrogram

```{r}
#| fig-height: 4
fit <- hclust(dist(sensory), method = "ward.D")
plot(fit, main = "", xlab = "pizza", sub = "", ylab = "merge height")
rect.hclust(fit, k = 3, border = pal[1:3])
```

::: small
Similarity is given by the **merge height**, not by how close the labels are on the x-axis.
:::

## k-means

```{r}
#| output-location: column
set.seed(1)
k3 <- kmeans(scale(sensory), centers = 3, nstart = 25)
k3$size
pc <- prcomp(sensory, scale. = TRUE)$x
plot(pc[, 1:2], pch = 19, cex = 1.6,
     col = pal[k3$cluster])
text(pc[, 1:2], labels = 1:16, pos = 3, cex = 0.7)
```

::: small
`nstart = 25` – the algorithm runs 25 times and keeps the best solution. Optimal k: `fviz_nbclust(..., method = "silhouette")`.
:::

## [QUIZ]{.quiz-label} Hierarchical or k-means?

You have 16 products and want to find out **how many** natural groups the market has. Which method do you start with?

. . .

::: answer
**Hierarchical clustering** – the dendrogram shows the whole structure and you choose the number of groups afterwards. k-means needs the number of clusters in advance.
:::

## Correspondence analysis

:::: columns
::: {.column width="45%"}
```{r}
age <- matrix(
  c(15, 2, 8, 7, 14, 4, 5, 5, 15, 11, 5, 9),
  ncol = 3, byrow = TRUE,
  dimnames = list(
    c("Men", "Women", "Children", "Seniors"),
    c("A", "B", "C")))
age
chisq.test(age)$p.value
```
:::

::: {.column width="55%"}
```{r}
#| echo: false
#| fig-width: 5.5
#| fig-height: 5
ca <- corresp(age, nf = 2)
plot(rbind(ca$rscore, ca$cscore), type = "n", asp = 1, xlab = "Dim 1", ylab = "Dim 2")
abline(h = 0, v = 0, lty = 3, col = "grey60")
text(ca$rscore, rownames(age), col = pal[3], font = 2)
text(ca$cscore, colnames(age), col = pal[2], font = 2, cex = 1.6)
```
:::
::::

::: small
A category in the same **direction from the centre** as a product chooses it more often than chance would suggest.
:::

## Summary

| question | method |
|---|---|
| are two attributes related? | correlation (Spearman), regression |
| which attributes go together? | PCA, factor analysis |
| which products are similar? | cluster analysis |
| which group chooses which product? | correspondence analysis |

::: links
[Theory 6](../../theory_EN/exercise06.html) [Theory 7](../../theory_EN/exercise07.html) [Theory 8](../../theory_EN/exercise08.html) [Theory 9](../../theory_EN/exercise09.html)
:::
