---
title: "SENSORY ANALYSIS: TWO-SAMPLE TESTS"
author: "A Sensometrics Tutorial"
format:
  revealjs:
    theme: simple
    slide-number: true
    chalkboard: true
    # Settings for robust scaling and overflow handling
    auto-scale: true
    scrollable: true
---

## Table of Contents

A complete exercise on statistical methods for comparing two samples in sensory evaluation.

-   **Part 1:** Discrimination Tests (Binomial Test)
-   **Part 2:** Paired Preference Test
-   **Part 3:** Continuous Data - Dependent Samples
-   **Part 4:** Continuous Data - Independent Samples
-   **Part 5:** McNemar Test (Before/After)
-   **Part 6:** Power Analysis
-   **Exercises & Solutions**

------------------------------------------------------------------------

## Theoretical Background

#### Types of Two-Sample Discrimination Tests:

1.  **Paired Comparison / Directional Difference Test**
    -   Panelists compare two samples on a specific attribute.
    -   Probability of guessing correctly: `p = 0.5` (50%)
2.  **Duo-Trio Test**
    -   One reference, two unknowns. Panelist identifies the unknown that matches the reference.
    -   Probability of guessing correctly: `p = 0.5` (50%)

------------------------------------------------------------------------

## Statistical Methods Overview

A. **For Discrete/Categorical Data:** - Binomial Test - McNemar Test - Chi-square Test

B. **For Continuous Data:** - **Parametric** (assumes normal distribution): - Paired t-test (dependent samples) - Independent t-test (independent samples) - **Non-parametric** (no normality assumption): - Wilcoxon Signed-Rank (dependent samples) - Mann-Whitney U (independent samples)

------------------------------------------------------------------------

## PART 1: Discrimination Tests (Binomial Test)

Used when data is binary (e.g., correct/incorrect, choice A/B).

**Example:** A paired comparison test asks "Which yogurt is creamier?" - 80 panelists participated. - 52 chose sample A. - Is this result significantly different from a 50/50 chance split?

```{r}
#| echo: true
#| output-location: fragment
# Perform the binomial test
binom.test(52, 80, p=0.5)
```

**Interpretation:** The p-value is 0.006. Since `p < 0.05`, we conclude that a significant number of panelists found yogurt A to be creamier.

------------------------------------------------------------------------

## PART 2: Paired Preference Test

Similar to discrimination, but focuses on preference.

**Example:** A preference test between an old and a new chocolate formulation. - 100 consumers participated. - 63 preferred the new formulation. - Is the new formulation significantly *preferred*? (one-sided test)

```{r}
#| echo: true
#| output-location: fragment
# Test if the new formulation is significantly better ("greater")
binom.test(63, 100, p=0.5, alternative="greater")
```

**Interpretation:** The p-value is 0.006. Since `p < 0.05`, we conclude that consumers have a statistically significant preference for the new chocolate formulation.

------------------------------------------------------------------------

## PART 3: Continuous Data - Dependent Samples

Used when the **same panelists** evaluate two products on a continuous scale.

**Example:** 20 panelists rated two beer samples on a 9-point hedonic scale.

**Steps:** 1. Check for normality (Shapiro-Wilk test, Q-Q plots). 2. Visualize the data (Boxplots). 3. Perform the appropriate statistical test: - If normal: **Paired t-test** - If not normal: **Wilcoxon Signed-Rank Test**

------------------------------------------------------------------------

### Dependent Samples: Data & Normality Check

```{r}
#| echo: true
# Data for two beers
beer_A = c(7.2, 6.5, 7.8, 6.9, 7.1, 6.8, 7.5, 7.0, 6.7, 7.3,
           6.9, 7.4, 7.1, 6.8, 7.2, 7.6, 6.9, 7.0, 7.3, 6.8)

beer_B = c(6.8, 6.2, 7.0, 6.5, 6.9, 6.4, 7.1, 6.7, 6.3, 6.9,
           6.5, 7.0, 6.8, 6.4, 6.9, 7.2, 6.5, 6.6, 6.8, 6.4)

# Shapiro-Wilk Normality Test
shapiro.test(beer_A)
shapiro.test(beer_B)
```

**Result:** Both p-values are \> 0.05, so we can assume the data is normally distributed. A Paired t-test is appropriate.

------------------------------------------------------------------------

### Dependent Samples: Visualization & Testing

::: {layout-ncol="2"}
#### Boxplot Comparison

```{r}
#| echo: true
#| fig-align: center
boxplot(beer_A, beer_B, 
        names=c("Beer A", "Beer B"),
        main="Beer Quality Comparison",
        ylab="Hedonic Score",
        col=c("lightblue", "lightgreen"))
points(1:2, c(mean(beer_A), mean(beer_B)), col="red", pch=19)
```

#### Paired t-test

```{r}
#| echo: true
# Perform the paired t-test
t.test(beer_A, beer_B, paired=TRUE)
```
:::

**Conclusion:** The p-value is `3.98e-06` (very small). We conclude that there is a highly significant difference in hedonic scores, with Beer A being rated higher than Beer B.

------------------------------------------------------------------------

## PART 4: Continuous Data - Independent Samples

Used when **two different groups of panelists** evaluate two products.

**Example:** - Group 1 (25 panelists) evaluated Cheese X. - Group 2 (30 panelists) evaluated Cheese Y.

**Steps:** 1. Check for normality (Shapiro-Wilk test for each group). 2. Check for homogeneity of variances (F-test). 3. Perform the appropriate statistical test: - If assumptions met: **Independent t-test** - If not: **Mann-Whitney U Test**

------------------------------------------------------------------------

### Independent Samples: Assumption Checks & Test

```{r}
#| echo: true
# Generate sample data
set.seed(123)
cheese_X = rnorm(25, mean=6.5, sd=1.2)
cheese_Y = rnorm(30, mean=7.1, sd=1.1)

# Check homogeneity of variances
var.test(cheese_X, cheese_Y)

# Perform Independent t-test
t.test(cheese_X, cheese_Y)
```

**Interpretation:** 1. **Variances:** The F-test p-value (0.64) is \> 0.05, so we can assume equal variances. 2. **t-test:** The p-value (0.09) is \> 0.05. We conclude there is no statistically significant difference in scores between Cheese X and Cheese Y.

------------------------------------------------------------------------

## PART 5: McNemar Test

Used for **paired categorical data**, often in "before and after" scenarios.

**Example:** 100 consumers stated their purchase intent ("Accept" vs. "Reject") for a product with old vs. new packaging. We want to know if the packaging change caused a significant shift in intent.

**Data:** - 45 accepted both. - 12 accepted the old but rejected the new. - 18 rejected the old but accepted the new. - 25 rejected both.

The test focuses on the "changers" (12 and 18).

------------------------------------------------------------------------

### McNemar Test: Application

```{r}
#| echo: true
# Create the contingency table
acceptance_data = matrix(c(45, 12, 18, 25), 
                        nrow=2,
                        dimnames=list(
                          "Old Formula" = c("Accept", "Reject"),
                          "New Formula" = c("Accept", "Reject")))

print(acceptance_data)

# Perform the McNemar test
mcnemar.test(acceptance_data, correct=FALSE)
```

**Interpretation:** The p-value (0.35) is \> 0.05. Although more people switched from rejecting to accepting (18) than the other way around (12), this difference is not statistically significant. The packaging change did not significantly affect purchase intent.

------------------------------------------------------------------------

## PART 6: Power Analysis

Helps determine the **sample size** needed to detect an effect of a certain size.

**Question:** How many panelists are needed in a paired comparison test to have an 80% chance (`power=0.80`) of detecting a medium effect size (`h=0.5`) at a significance level of 0.05?

```{r}
#| echo: true
#| warning: false
# Install and load the pwr package
if(!require(pwr)) install.packages("pwr")
library(pwr)

# Calculate sample size
pwr.p.test(h=0.5, sig.level=0.05, power=0.80)
```

**Conclusion:** You would need approximately **32 panelists** to reliably detect a medium-sized effect.

------------------------------------------------------------------------

## EXERCISES

Time to apply what we've learned!

------------------------------------------------------------------------

### EXERCISE 1: Duo-Trio Test

A food company reformulated their orange juice. In a duo-trio test: - 120 panelists participated. - 71 correctly identified which sample matched the reference.

**Tasks:** a) Perform the appropriate statistical test. b) Is there a significant difference between the old and new formulation?

------------------------------------------------------------------------

### SOLUTION 1: Duo-Trio Test

The appropriate test is the **Binomial Test**, with a null probability of 0.5.

```{r}
#| echo: true
# a) Statistical test
result1 = binom.test(71, 120, p=0.5)
print(result1)
```

**b) Interpretation:** The p-value is **0.033**. Since `p < 0.05`, we conclude that **yes, there is a significant difference**. Panelists were able to reliably detect a difference between the two formulations.

------------------------------------------------------------------------

### EXERCISE 3: Paired Hedonic Scores

The same 15 panelists evaluated two wine samples (conventional vs. organic) on a 9-point hedonic scale. Is the organic wine significantly preferred?

```{r}
#| echo: true
#| eval: false
wine_conventional = c(7.5, 6.8, 7.2, 6.9, 7.8, 7.1, 6.5, 7.4, 
                     7.0, 6.9, 7.3, 7.6, 6.8, 7.2, 7.1)
wine_organic = c(7.8, 7.2, 7.5, 7.3, 8.1, 7.4, 6.9, 7.7, 
                7.3, 7.2, 7.6, 7.9, 7.1, 7.5, 7.4)
```

**Tasks:** a) Check normality of the differences. b) Perform the appropriate paired test. c) Interpret the results.

------------------------------------------------------------------------

### SOLUTION 3: Paired Hedonic Scores

Since the same panelists rated both wines, this is a **dependent samples** problem.

```{r}
#| echo: true
wine_conventional = c(7.5, 6.8, 7.2, 6.9, 7.8, 7.1, 6.5, 7.4, 7.0, 6.9, 7.3, 7.6, 6.8, 7.2, 7.1)
wine_organic = c(7.8, 7.2, 7.5, 7.3, 8.1, 7.4, 6.9, 7.7, 7.3, 7.2, 7.6, 7.9, 7.1, 7.5, 7.4)

# a) Check normality of differences
shapiro.test(wine_organic - wine_conventional)

# b) Perform paired t-test (since data is normal)
t.test(wine_organic, wine_conventional, paired=TRUE)
```

**c) Interpretation:** The p-value is **0.00015**. This is highly significant. We conclude that **consumers significantly prefer the organic wine** over the conventional wine.

------------------------------------------------------------------------

## Summary: Choosing the Right Test {.smaller}

| Situation | Data Type | Same Panelists? | Normally Distributed? | Recommended Test |
|---------------|---------------|---------------|---------------|---------------|
| Discrimination Task | Binary | N/A | N/A | **Binomial Test** |
| Before/After Acceptance | Binary | Yes | N/A | **McNemar Test** |
| Hedonic/Intensity Scores | Continuous | Yes | Yes | **Paired t-test** |
| Hedonic/Intensity Scores | Continuous | Yes | No | **Wilcoxon Signed-Rank** |
| Hedonic/Intensity Scores | Continuous | No | Yes | **Independent t-test** |
| Hedonic/Intensity Scores | Continuous | No | No | **Mann-Whitney U** |

------------------------------------------------------------------------

## Key Concepts to Remember

1.  **Check Your Assumptions:** Always test for normality (and equal variances for independent tests) before choosing a parametric test.

2.  **Paired vs. Independent:** This is the most critical distinction. Is it the *same group* of people or *different groups*?

3.  **p-value \< 0.05 is not everything:** A significant result tells you there is a difference, but not how *large* or *meaningful* it is
